Graph embedding and transfer learning can help predict potential species interaction networks despite data limitations

### 2023-09-02 index.qmd
ready

---
title: Graph embedding and transfer learning can help predict potential species interaction networks despite data limitations
authors:
  - name: Tanya Strydom
    affiliation:
        - ref: udem
        - ref: qcbs
    orcid: 0000-0001-6067-1349
    email: tanya.strydom@umontreal.ca
  - name: Salomé Bouskila
    affiliation:
        - ref: udem
    roles: writing
    corresponding: false
    orcid: 0000-0003-0193-5441
    email: salome.bouskila@umontreal.ca
  - name: Francis Banville
    affiliation:
        - ref: udem
        - ref: qcbs
        - ref: udes
    orcid: 0000-0001-9051-0597
    email: francis.banville@umontreal.ca
  - name: Ceres Barros
    affiliation:
        - ref: ubcforest
    orcid: 0000-0003-4036-977X
    email: cbarros@mail.ubc.ca
  - name: Dominique Caron
    affiliation:
        - ref: mcgill
        - ref: qcbs
    orcid: 0000-0002-2151-6693
    email: dominique.caron@mail.mcgill.ca
  - name: Maxwell J. Farrell
    affiliation:
        - ref: uoft
    orcid: 0000-0003-0452-6993
    email: maxwell.farrell@utoronto.ca
  - name: Marie-Josée Fortin
    affiliation:
        - ref: uoft
    orcid: 0000-0002-9935-1366
    email: mariejosee.fortin@utoronto.ca
  - name: Benjamin Mercier
    affiliation:
        - ref: udes
        - ref: qcbs
    orcid: 0000-0002-4104-9463
    email: benjamin.b.mercier@usherbrooke.ca
  - name: Laura J. Pollock
    affiliation:
        - ref: mcgill
        - ref: qcbs
    orcid: 0000-0002-6004-4027
    email: laura.pollock@mcgill.ca
  - name: Rogini Runghen
    affiliation:
        - ref: ucecology
    orcid: 0000-0001-5785-8321
    email: rogini.runghen@gmail.com
  - name: Giulio V. Dalla Riva
    affiliation:
        - ref: ucmath
    orcid: 0000-0002-3454-0633
    email: giulio.dallariva@canterbury.ac.nz
  - name: Timothée Poisot
    affiliation:
        - ref: udem
        - ref: qcbs
    orcid: 0000-0002-0735-5184
    corresponding: true
    email: timothee.poisot@umontreal.ca
affiliations:
    - id: udem
      name: Département de Sciences Biologiques, Université de Montréal, Montréal, Canada
    - id: qcbs
      name: Québec Centre for Biodiversity Science, Montréal, Canada
    - id: udes
      name: Département de Biologie, Université de Sherbrooke, Sherbrooke, Canada
    - id: ubcforest
      name: Department of Forest Resources Management, University of British Columbia, Vancouver, B.C., Canada
    - id: mcgill
      name: Department of Biology, McGill University, Montréal, Canada
    - id: uoft
      name: Department of Ecology & Evolutionary Biology, University of Toronto, Toronto, Canada
    - id: ubccons
      name: Department of Forest and Conservation Sciences, University of British Columbia, Vancouver, Canada
    - id: ucecology
      name: Department of Forest and Conservation Sciences, University of British Columbia, Vancouver, Canada
    - id: ucmath
      name: Department of Forest and Conservation Sciences, University of British Columbia, Vancouver, Canada
bibliography: references.bib
---

The ability to infer *potential* interactions could serve as a significant breakthrough in our ability to conceptualize species interaction networks over large spatial scales [@Hortal2015Seven]. Reliable inferences would not only boost our understanding of the structure of species interaction networks, but also increase the amount of information that can be used for biodiversity management. In a recent overview of the field of ecological network prediction, @Strydom2021Roadmap identified two challenges of interest to the prediction of interactions at large scales. First, there is a relative scarcity of relevant data in most places globally -- which, due to the limitations in most predictive methods, restricts the ability to infer interactions to locations where it is least required (*i.e.* regions where we already have interaction data) leaving us unable to make inference in data scarce regions (where we most need it); second, accurate predictors are important for accurate predictions, and the lack of methods that can leverage a small amount of *accurate* data is a serious impediment to our predictive ability. In this contribution, we highlight the power of viewing (and constructing) metawebs as *probabilistic* objects in the context of low-probability interactions, discuss how a family of machine learning tools (graph embeddings and transfer learning) can be used to overcome data limitations to metaweb inference, and highlight how the use of metawebs introduces important questions for the field of network ecology.

In most places, our most reliable biodiversity knowledge is that of a species pool where a set of potentially interacting species in a given area could occur: through the analysis of databases like the Global Biodiversity Information Facility (GBIF) or the International Union for the Conservation of Nature (IUCN), it is possible to construct a list of species for a region of interest. Following the definition of @Dunne2006Network, a metaweb is the ecological network analogue to the species pool; specifically, it inventories all *potential* interactions between species for a spatially delimited area (and so captures the $\gamma$ diversity of interactions as per @Poisot2012Dissimilarity). However, inferring the potential interactions between these species still remains a challenge. And yet, because the metaweb represents the joint effect of functional, phylogenetic, and macroecological processes [@Morales-Castilla2015Inferring; @carlson2022; @morales-castilla2021], it holds valuable ecological information. Specifically, it represents the "upper bounds" on what the composition of the local networks, given a local species pool, can be [see *e.g.* @McLeod2021Sampling]; this information can help evaluate the ability of ecological assemblages to withstand the effects of, for example, climate change [@Fricke2022Effects]. These local networks may be reconstructed given an appropriate knowledge of local species composition and provide information on the structure of networks at finer spatial scales. This has been done for example for tree-galler-parasitoid systems [@Gravel2018Bringing], fish trophic interactions [@Albouy2019Marine], tetrapod trophic interactions [@Braga2019Spatial; @OConnor2020Unveiling], and crop-pest networks [@Grunig2020Crop].

The metaweb itself is not a prediction of local networks at specific locations within the spatial area it covers: it will have a different structure, notably by having a larger connectance [see *e.g.* @Wood2015Effects] and complexity [see *e.g.* @Galiana2022Ecological], than any of these local networks. These local networks (which capture the $\alpha$ diversity of interactions) are a subset of the metaweb's species and its realized interactions, and have been called "metaweb realizations" [@Poisot2015Species]. Differences between local networks and their metawebs are due to chance, species abundance and co-occurrence, local environmental conditions, and local distribution of functional traits, among others. Specifically, although co-occurrence can be driven by interactions [@Cazelles2016Theory], co-occurrence alone is not a predictor of interactions [@Blanchet2020Cooccurrence; @Thurman2019Testing], and therefore the lack of co-occurrence cannot be used to infer the lack of a feasible interaction. Yet, recent results by @Saravia2021Ecological strongly suggested that local (metaweb) realizations only respond weakly to local conditions: instead, they reflect constraints inherited by the structure of their metaweb. This sets up the core goal of predictive network ecology as the prediction of metaweb structure, as it is required to accurately produce downscaled, local predictions.

@@

Treating interactions as probabilistic (as opposed to binary) events is a more nuanced and realistic way to represent them. @Dallas2017Predicting suggested that most interactions (links) in ecological networks are cryptic, *i.e.* uncommon or hard to observe. This argument echoes @Jordano2016Sampling: sampling ecological interactions is difficult because it requires first the joint observation of two species, and then the observation of their interaction. In addition, it is generally expected that weak or rare interactions will be more prevalent in networks than common or strong interactions [@Csermely2004Strong], compared to strong, persistent interactions; this is notably the case in food chains, wherein many weaker interactions are key to the stability of a system [@Neutel2002Stability]. In the light of these observations, we expect to see an over-representation of low-probability (hereafter rare) interactions under a model that accurately predicts interaction probabilities.

Yet, the original metaweb definition, and indeed most past uses of metawebs, was based on the presence/absence of interactions. Moving towards *probabilistic* metawebs, by representing interactions as Bernoulli events [see *e.g.* @Poisot2016Structure], offers the opportunity to weigh these rare interactions appropriately. The inherent plasticity of interactions is important to capture: there have been documented instances of food webs undergoing rapid collapse/recovery cycles over short periods of time [*e.g.* @Pedersen2017Signatures]. Furthermore, because the structure of the metaweb cannot be known in advance, it is important to rely on predictive tools that do not assume a specific network topology for link prediction [@Gaucher2021Outlier], but are able to work on generalizations of the network. These considerations emphasize why metaweb predictions should focus on quantitative (preferentially probabilistic) predictions, and this should constrain the suite of models that are appropriate for prediction. Binary classifiers based on probabilities have an extremely robust methodology to validate them, and this applies naturally to the prediction of interactions [@poisot2023].

It is important to recall that a metaweb is intended as a catalogue of all potential (feasible) interactions, which is then filtered for a given application [@Morales-Castilla2015Inferring]. It is therefore important to separate the interactions that happen "almost surely" (repeated observational data), "almost never" (repeated lack of evidence *or* evidence that the link is forbidden through *e.g.* trait mis-match), and interactions with a probability that lays somewhere in between [@Catchen2023Missing]. Although metawebs can (and in practice likely do) include false positives, these are statistically negligible compared to the false negatives. Furthermore, @Strydom2022Food shows that t-SVD embedding is extremely robust to (and able to detect) the presence of false positives. In a sense, that most ecological interactions are elusive can call for a slightly different approach to sampling: once the common interactions are documented, the effort required in documenting each rare interaction will increase exponentially [@Jordano2016Sampling]. Recent proposals in other fields relying on machine learning approaches emphasize the idea that algorithms meant to predict, through the assumption that they approximate the process generating the data, can also act as data generators [@Hoffmann2019Machine]. High quality observational data can be used to infer core rules underpinning network structure, and be supplemented with synthetic data coming from predictive models trained on them, thereby increasing the volume of information available for analysis. Indeed, @Strydom2021Roadmap suggested that knowing the metaweb may render the prediction of local networks easier, because it fixes an "upper bound" on which interactions can exist. In this context, a probabilistic metaweb represents an aggregation of informative priors on the biological feasibility of interactions, which is usually hard to obtain yet has possibly the most potential to boost our predictive ability of local networks [@Bartomeus2013Understanding; @Bartomeus2016Common]. This would represent a departure from simple rules expressed at the network scale [*e.g.* @Williams2000Simple] to a view of network prediction based on learning the rules that underpin interactions *and* their variability [@Gupta2022Simultaneously].

![The embedding process (**A**) can help to identify links (interactions) that may have been missed within the original community (represented by the orange dashed arrows, **B**). Transfer learning (**D**) allows for the prediction links (interactions) even when novel species (**C**) are included alongside the original community. This is achieved by learning using other relevant predictors (*e.g.* traits) in conjunction with the known interactions to infer latent values (**E**). Ultimately this allows us to predict links (interactions) for species external from the original sample (blue dashed arrows) as well as missing within sample links (**F**). Within this context the predicted (and original) networks as well as the ecological predictors used (green boxes) are products that can be quantified through measurements in the field, whereas the embedded as well as imputed matrices (purple box) are representative of a decomposition of the interaction matrices onto the embedding space](figures/conceptual_2.png){#fig-embedding}

@@

| Method        | Object          | Technique                        | Reference                | Application                                                                                                       |
|---------------|---------------|---------------|---------------|---------------|
| tSNE          | nodes           | statistical divergence           | @Hinton2002Stochastic    | [@Cieslak2020Tdistributed, species-environment responses $^a$] [@Gibb2021Data, host-virus network representation] |
| LINE          | nodes           | stochastic gradient descent      | @Tang2015Line            |                                                                                                                   |
| SDNE          | nodes           | gradient descent                 | @Wang2016Structural      |                                                                                                                   |
| node2vec      | nodes           | stochastic gradient descent      | @Grover2016Node2vec      |                                                                                                                   |
| HARP          | nodes           | meta-strategy                    | @Chen2017Harp            |                                                                                                                   |
| DMSE          | joint nodes     | deep neural network              | @Chen2017Deep            | [@Chen2017Deep, species-environment interactions $^b$]                                                            |
| graph2vec     | sub-graph       | skipgram network                 | @Narayanan2017Graph2vec  |                                                                                                                   |
| RDPG          | graph           | SVD                              | @Young2007Random         | [@DallaRiva2016Exploring, trophic interactions] [@Poisot2021Imputing, host-virus network prediction]              |
| GLEE          | graph           | Laplacian eigenmap               | @Torres2020Glee          |                                                                                                                   |
| DeepWalk      | graph           | stochastic gradient descent      | @Perozzi2014Deepwalk     | [@Wardeh2021Predicting, host-virus interactions]                                                                  |
| GraphKKE      | graph           | stochastic differential equation | @Melnyk2020Graphkke      | [@Melnyk2020Graphkke, microbiome species associations $^a$]                                                       |
| FastEmbed     | graph           | eigen decomposition              | @Ramasamy2015Compressive |                                                                                                                   |
| PCA           | graph           | eigen decomposition              | @Surendran2013Graph      | [@Strydom2021Roadmap, host-parasite interactions]                                                                 |
| Joint methods | multiple graphs | multiple strategies              | @Wang2021Joint           |                                                                                                                   |

: Overview of some common graph embedding approaches, by type of embedded objects, alongside examples of their use in the prediction of species interactions. These methods have not yet been routinely used to predict species interactions; most examples that we identified were either statistical associations, or analogues to joint species distribution models. $^a$: application is concerned with *statistical* interactions, which are not necessarilly direct biotic interactions; $^b$:application is concerned with joint-SDM-like approach, which is also very close to statistical associations as opposed to direct biotic interactions. Given the need to evaluate different methods on a problem-specific basis, the fact that a lot of methods have not been used on network problems is an opportunity for benchmarking and method development. Note that the row for PCA also applies to kernel/probabilistic PCA, which are variations on the more general method of SVD. Note further that tSNE has been included because it is frequently used to embed graphs, including of species associations/interactions, despite not being strictly speaking, a graph embedding technique [see *e.g.* @Chami2022Machine]. {#tbl-methods}

@@

In @fig-illustration-1, we focus on some statistical checks of the embedding. In panel **A**, we show that the averaged $L_2$ loss (*i.e.* the mean of squared errors) between the empirical and reconstructed metaweb decreases when the number of dimensions (rank) of the subspace increases, with an inflection at 39 dimensions (out of 120 initially) according to the finite differences method. As discussed by @Runghen2021Exploiting, there is often a trade-off between the number of dimensions to use (more dimensions are more computationally demanding) and the quality of the representation. In panel **B**, we show the increase in cumulative variance explained at each rank, and visualize that using 39 ranks explains about 70% of the variance in the empirical metaweb. This is a different information from the $L_2$ loss (which is averaged across interactions), as it works on the eigenvalues of the embedding, and therefore captures higher-level features of the network. In panel **C**, we show positions of hosts and parasites on the first two dimensions of the left and right subspaces. Note that these values largely skew negative, because the first dimensions capture the coarse structure of the network: most pairs of species do not interact, and therefore have negative values. Finally in panel **D**, we show the predicted weight (*i.e.* the result of the multiplication of the RDGP subspaces at a rank of 39) as a function of whether the interactions are observed, not-observed, or unknown due to lack of co-occurrence in the original dataset. This reveals that the observed interactions have higher predicted weights, although there is some overlap; the usual approach to identify potential interactions based on this information would be a thresholding analysis, which is outside the scope of this manuscript (and is done in the papers cited in this illustration). Because the values returned from RDPG are not bound to the unit interval, we performed a clamping of the weights to the unit space, showing a one-inflation in documented interactions, and a zero-inflation in other species pairs. This last figure shows that species pairs with no documented co-occurrence have weights that are not distinguishable from species pairs with no documented interactions; in other words, looking at the embedding, species that do not co-occur are not easily distinguished from species that do not interact. This suggests that (as befits a host-parasite model) the ability to interact is a strong predictor of co-occurrence.

@@

The results of @fig-illustration-1 show that we can extract an embedding of the metaweb that captures enough variance to be relevant; specifically, this is true for both $L_2$ loss (indicating that RDPG is able to capture pairwise processes) and the cumulative variance explained (indicating that RDPG is able to capture network-level structure). Therefore, in @fig-illustration-2, we relate the values of latent variables for hosts to different ecologically-relevant data. In panel **A**, we show that host with a higher value on the first dimension have fewer parasites. This relates to the body size of hosts in the *PanTHERIA* database [@Jones2009Pantheria], as shown in panel **B**: interestingly, the position on the first axis is only weakly correlated to body mass of the host; this matches well established results showing that body size/mass is not always a direct predictor of parasite richness in terrestrial mammals [@Morand1998Density], a result we observe in panel **C**. Finally, in panel **D**, we can see how different taxonomic families occupy different positions on the first axis, with *e.g.* Sciuridae being biased towards higher values. These results show how we can look for ecological informations in the output of network embeddings, which can further be refined into the selection of predictors for transfer learning.

@@

Metaweb inference seeks to provide information about the interactions between species at a large spatial scale, typically a scale large enough to be considered of biogeographic relevance (indeed, many of the examples covered in the introduction span areas larger than a country, some of them global). But as @Herbert1965Dune rightfully pointed out, "\[y\]ou can't draw neat lines around planet-wide problems"; any inference of a metaweb must therefore contend with several novel, interwoven, families of problems. In this section, we outline three that we think are particularly important, and can discuss how they may addressed with subsequent data analysis or simulations, and how they emerge in the specific context of using embeddings; some of these issues are related to the application of these methods at the science-policy interface. Adressing these considerations as part of the methodological discussion is particularly important, as the construction of metawebs can perpetuate legacies of biases in data (Box 2).

>>>>

---
title: Graph embedding and transfer learning can help predict potential species interaction networks despite data limitations
authors:
  - name: Tanya Strydom
    affiliation:
        - ref: udem
        - ref: qcbs
    orcid: 0000-0001-6067-1349
    email: tanya.strydom@umontreal.ca
  - name: Salomé Bouskila
    affiliation:
        - ref: udem
    roles: writing
    corresponding: false
    orcid: 0000-0003-0193-5441
    email: salome.bouskila@umontreal.ca
  - name: Francis Banville
    affiliation:
        - ref: udem
        - ref: qcbs
        - ref: udes
    orcid: 0000-0001-9051-0597
    email: francis.banville@umontreal.ca
  - name: Ceres Barros
    affiliation:
        - ref: ubcforest
    orcid: 0000-0003-4036-977X
    email: cbarros@mail.ubc.ca
  - name: Dominique Caron
    affiliation:
        - ref: mcgill
        - ref: qcbs
    orcid: 0000-0002-2151-6693
    email: dominique.caron@mail.mcgill.ca
  - name: Maxwell J. Farrell
    affiliation:
        - ref: uoft
    orcid: 0000-0003-0452-6993
    email: maxwell.farrell@utoronto.ca
  - name: Marie-Josée Fortin
    affiliation:
        - ref: uoft
    orcid: 0000-0002-9935-1366
    email: mariejosee.fortin@utoronto.ca
  - name: Benjamin Mercier
    affiliation:
        - ref: udes
        - ref: qcbs
    orcid: 0000-0002-4104-9463
    email: benjamin.b.mercier@usherbrooke.ca
  - name: Laura J. Pollock
    affiliation:
        - ref: mcgill
        - ref: qcbs
    orcid: 0000-0002-6004-4027
    email: laura.pollock@mcgill.ca
  - name: Rogini Runghen
    affiliation:
        - ref: ucecology
    orcid: 0000-0001-5785-8321
    email: rogini.runghen@gmail.com
  - name: Giulio V. Dalla Riva
    affiliation:
        - ref: ucmath
    orcid: 0000-0002-3454-0633
    email: giulio.dallariva@canterbury.ac.nz
  - name: Timothée Poisot
    affiliation:
        - ref: udem
        - ref: qcbs
    orcid: 0000-0002-0735-5184
    corresponding: true
    email: timothee.poisot@umontreal.ca
affiliations:
    - id: udem
      name: Département de Sciences Biologiques, Université de Montréal, Montréal, Canada
    - id: qcbs
      name: Québec Centre for Biodiversity Science, Montréal, Canada
    - id: udes
      name: Département de Biologie, Université de Sherbrooke, Sherbrooke, Canada
    - id: ubcforest
      name: Department of Forest Resources Management, University of British Columbia, Vancouver, B.C., Canada
    - id: mcgill
      name: Department of Biology, McGill University, Montréal, Canada
    - id: uoft
      name: Department of Ecology & Evolutionary Biology, University of Toronto, Toronto, Canada
    - id: ubccons
      name: Department of Forest and Conservation Sciences, University of British Columbia, Vancouver, Canada
    - id: ucecology
      name: Department of Forest and Conservation Sciences, University of British Columbia, Vancouver, Canada
    - id: ucmath
      name: Department of Forest and Conservation Sciences, University of British Columbia, Vancouver, Canada
bibliography: references.bib
abstract: |
    - Metawebs (networks of potential interactions within a species pool) are a powerful abstraction to understand how large-scale species interaction networks are structured.
    - Because metawebs are typically expressed at large spatial and taxonomic scales, assembling them is a tedious and costly process; predictive methods can help circumvent the limitations in data deficiencies, by providing a first approximation of metawebs."
    - One way to improve our ability to predict metawebs is to maximize available information by using graph embeddings, as opposed to an exhaustive list of species interactions. Graph embedding is an emerging field in machine learning that holds great potential for ecological problems.
    - Here, we outline how the challenges associated with inferring metawebs line-up with the advantages of graph embeddings; followed by a discussion as to how the choice of the species pool has consequences on the reconstructed network, specifically as to the role of human-made (or arbitrarily assigned) boundaries and how these may influence ecological hypotheses.
---

The ability to infer potential biotic interactions could serve as a significant breakthrough in our ability to conceptualize networks over large spatial scales [@Hortal2015Seven]. Reliable inferences would not only boost our understanding of the structure of species interaction networks, but also increase the amount of information that can be used for biodiversity management. In a recent overview of the field of ecological network prediction, @Strydom2021Roadmap identified two challenges of interest to the prediction of interactions at large scales. First, there is a relative scarcity of relevant data in most places globally -- which, due to the limitations in most predictive methods, restricts the ability to infer interactions to locations where it is least required (*i.e.* regions where we already have interaction data) leaving us unable to make inference in data scarce regions (where we most need it); second, accurate predictors are important for accurate predictions, and the lack of methods that can leverage a small amount of *accurate* data is a serious impediment to our predictive ability. In this contribution, we (i) highlight the power of viewing (and constructing) metawebs as *probabilistic* objects in the context of low-probability interactions, (ii) discuss how a family of machine learning tools (graph embeddings and transfer learning) can be used to overcome data limitations to metaweb inference, and (iii) highlight how the use of metawebs introduces important questions for the field of network ecology.

In most places, our most reliable biodiversity knowledge is that of a species pool where a set of potentially interacting species in a given area could occur: through the analysis of databases like the Global Biodiversity Information Facility (GBIF) or the International Union for the Conservation of Nature (IUCN), it is possible to construct a list of species for a region of interest. Following the definition of @Dunne2006Network, a metaweb is the ecological network analogue to the species pool; specifically, it inventories all *potential* interactions between species for a spatially delimited area (and so captures the $\gamma$ diversity of interactions as per @Poisot2012Dissimilarity). However, inferring the potential interactions between these species still remains a challenge. And yet, the metaweb holds valuable ecological information: it represents the joint effect of functional, phylogenetic, and macroecological processes [@Morales-Castilla2015Inferring; @carlson2022; @morales-castilla2021]. Specifically, it represents the "upper bounds" on what the composition of the local networks, given a local species pool, can be [see *e.g.* @McLeod2021Sampling]; this information can help evaluate the ability of ecological assemblages to withstand the effects of, for example, climate change [@Fricke2022Effects]. These local networks may be reconstructed given an appropriate knowledge of local species composition and provide information on the structure of networks at finer spatial scales. This has been done for example for tree-galler-parasitoid systems [@Gravel2018Bringing], fish trophic interactions [@Albouy2019Marine], terrestrial tetrapod trophic interactions [@Braga2019Spatial; @OConnor2020Unveiling], and crop-pest networks [@Grunig2020Crop].

The metaweb itself is not a prediction of local networks at specific locations within the spatial area it covers: it will have a different structure, notably by having a larger connectance [see *e.g.* @Wood2015Effects] and complexity [see *e.g.* @Galiana2022Ecological], than any of these local networks. Local networks (which capture the $\alpha$ diversity of interactions) are a subset of the metaweb's species and its realized interactions, and have been called "metaweb realizations" [@Poisot2015Species]. Differences between local networks and their metawebs are due to chance, species abundance and co-occurrence, local environmental conditions, and local distribution of functional traits, among others. Specifically, although co-occurrence can be driven by interactions [@Cazelles2016Theory], co-occurrence alone is not a predictor of interactions [@Blanchet2020Cooccurrence; @Thurman2019Testing], and therefore the lack of co-occurrence cannot be used to infer the lack of a feasible interaction. Yet, recent results by @Saravia2021Ecological strongly suggested that local (metaweb) realizations only respond weakly to local conditions: instead, they reflect constraints inherited by the structure of their metaweb. This sets up the core goal of predictive network ecology as the prediction of metaweb structure, as it is required to accurately produce downscaled, local predictions.

@@

Treating interactions as probabilistic (as opposed to binary) events is a more nuanced and realistic way to represent them. @Dallas2017Predicting suggested that most interactions (links) in ecological networks are cryptic, *i.e.* uncommon or hard to observe. This argument echoes @Jordano2016Sampling: sampling ecological interactions is difficult because it requires first the joint observation of two species, and then the observation of their interaction. In addition, it is generally expected that weak or rare interactions will be more prevalent in networks than common or strong interactions [@Csermely2004Strong]; this is notably the case in food chains, wherein many weaker interactions are key to the stability of a system [@Neutel2002Stability]. In the light of these observations, we expect to see an over-representation of low-probability (hereafter rare) interactions under a model that accurately predicts interaction probabilities.

Yet, the original metaweb definition, and indeed most past uses of metawebs, was based on the presence/absence of interactions. Moving towards *probabilistic* metawebs, by representing interactions as Bernoulli events [see *e.g.* @Poisot2016Structure], offers the opportunity to weigh these rare interactions appropriately. The inherent plasticity of interactions is important to capture: there have been documented instances of food webs undergoing rapid collapse/recovery cycles over short periods of time [*e.g.* @Pedersen2017Signatures]. Furthermore, because the structure of the metaweb cannot be known in advance, it is important to rely on predictive tools that do not assume a specific network topology for link prediction [@Gaucher2021Outlier], but are able to work on generalizations of the network that capture statistical processes giving it its structure. These considerations emphasize why metaweb predictions should focus on quantitative (preferentially probabilistic) predictions, and this should constrain the suite of models that are appropriate for prediction. Binary classifiers based on probabilities have an extremely robust methodology to validate them, and this applies naturally to the prediction of interactions [@poisot2023].

It is important to recall that a metaweb is intended as a catalogue of all potential (feasible) interactions, which is then filtered for a given application [@Morales-Castilla2015Inferring]. It is therefore important to separate the interactions that happen "almost surely" (repeated observational data), "almost never" (repeated lack of evidence *or* evidence that the link is forbidden through *e.g.* trait mis-match), and interactions with a probability that lays somewhere in between [@Catchen2023Missing]. Although metawebs can (and in practice likely do) include false positives, these are statistically negligible compared to the false negatives. Furthermore, @Strydom2022Food shows that t-SVD embedding is extremely robust to (and able to detect) the presence of false positives. In a sense, because most ecological interactions are elusive, we should consider the direct consequences this has on sampling: once the common interactions are documented, the effort required in documenting each rare interaction will increase exponentially [@Jordano2016Sampling]. Recent proposals in other fields relying on machine learning approaches emphasize the idea that algorithms meant to predict, through the assumption that they approximate the process generating the data, can also act as data generators [@Hoffmann2019Machine]. High quality observational data can be used to infer core rules underpinning network structure, and be supplemented with synthetic data coming from predictive models trained on them, thereby increasing the volume of information available for analysis. Indeed, @Strydom2021Roadmap suggested that knowing the metaweb may render the prediction of local networks easier, because it fixes an "upper bound" on which interactions can exist. In this context, a probabilistic metaweb represents an aggregation of informative priors on the biological feasibility of interactions, which is usually hard to obtain yet has possibly the most potential to boost our predictive ability of local networks [@Bartomeus2013Understanding; @Bartomeus2016Common]. This would represent a departure from simple rules expressed at the network scale [*e.g.* @Williams2000Simple] to a view of network prediction based on learning the rules that underpin interactions *and* their variability [@Gupta2022Simultaneously].

![The embedding process (**A**) can help to identify links (interactions) that may have been missed within the original community (represented by the orange dashed arrows, **B**). Transfer learning (**D**) allows for the prediction of links (interactions) even when novel species (**C**) are included alongside the original community. This is achieved with other ecologically relevant predictors (*e.g.* traits) in conjunction with the known interactions to infer latent values (**E**). Ultimately this allows us to predict links (interactions) for species external from the original sample (blue dashed arrows) as well as missing within sample links (**F**). Within this context the predicted (and original) networks as well as the ecological predictors used (green boxes) are products that can be quantified through measurements in the field, whereas the embedded as well as imputed matrices (purple box) are representative of a decomposition of the interaction matrices onto the embedding space]

  [The embedding process (**A**) can help to identify links (interactions) that may have been missed within the original community (represented by the orange dashed arrows, **B**). Transfer learning (**D**) allows for the prediction of links (interactions) even when novel species (**C**) are included alongside the original community. This is achieved with other ecologically relevant predictors (*e.g.* traits) in conjunction with the known interactions to infer latent values (**E**). Ultimately this allows us to predict links (interactions) for species external from the original sample (blue dashed arrows) as well as missing within sample links (**F**). Within this context the predicted (and original) networks as well as the ecological predictors used (green boxes) are products that can be quantified through measurements in the field, whereas the embedded as well as imputed matrices (purple box) are representative of a decomposition of the interaction matrices onto the embedding space]: figures/conceptual_2.png {#fig-embedding}

@@

| Method        | Object          | Technique                        | Reference                | Application                                                                                                       |
|---------------|---------------|---------------|---------------|---------------|
| t-SNE         | nodes           | statistical divergence           | @Hinton2002Stochastic    | [@Cieslak2020Tdistributed, species-environment responses $^a$] [@Gibb2021Data, host-virus network representation] |
| LINE          | nodes           | stochastic gradient descent      | @Tang2015Line            |                                                                                                                   |
| SDNE          | nodes           | gradient descent                 | @Wang2016Structural      |                                                                                                                   |
| node2vec      | nodes           | stochastic gradient descent      | @Grover2016Node2vec      |                                                                                                                   |
| HARP          | nodes           | meta-strategy                    | @Chen2017Harp            |                                                                                                                   |
| DMSE          | joint nodes     | deep neural network              | @Chen2017Deep            | [@Chen2017Deep, species-environment interactions $^b$]                                                            |
| graph2vec     | sub-graph       | skipgram network                 | @Narayanan2017Graph2vec  |                                                                                                                   |
| RDPG          | graph           | SVD                              | @Young2007Random         | [@DallaRiva2016Exploring, trophic interactions] [@Poisot2021Imputing, host-virus network prediction]              |
| GLEE          | graph           | Laplacian eigenmap               | @Torres2020Glee          |                                                                                                                   |
| DeepWalk      | graph           | stochastic gradient descent      | @Perozzi2014Deepwalk     | [@Wardeh2021Predicting, host-virus interactions]                                                                  |
| GraphKKE      | graph           | stochastic differential equation | @Melnyk2020Graphkke      | [@Melnyk2020Graphkke, microbiome species associations $^a$]                                                       |
| FastEmbed     | graph           | eigen decomposition              | @Ramasamy2015Compressive |                                                                                                                   |
| PCA           | graph           | eigen decomposition              | @Surendran2013Graph      | [@Strydom2021Roadmap, host-parasite interactions]                                                                 |
| Joint methods | multiple graphs | multiple strategies              | @Wang2021Joint           |                                                                                                                   |

: Overview of some common graph embedding approaches, by type of embedded objects, alongside examples of their use in the prediction of species interactions. These methods have not yet been routinely used to predict species interactions; most examples that we identified were either statistical associations, or analogues to joint species distribution models. $^a$: application is concerned with *statistical* interactions, which are not necessarilly direct biotic interactions; $^b$:application is concerned with joint-SDM-like approach, which is also very close to statistical associations as opposed to direct biotic interactions. Given the need to evaluate different methods on a problem-specific basis, the fact that many methods have not been used on network problems is an opportunity for benchmarking and method development. Note that the row for PCA also applies to kernel/probabilistic PCA, which are variations on the more general method of SVD. Note further that t-SNE has been included because it is frequently used to embed graphs, including of species associations/interactions, despite not being strictly speaking, a graph embedding technique [see *e.g.* @Chami2022Machine]. {#tbl-methods}

@@

In @fig-illustration-1, we focus on some statistical checks of the embedding. In panel **A**, we show that the averaged $L_2$ loss (*i.e.* the mean of squared errors) between the empirical and reconstructed metaweb decreases when the number of dimensions (rank) of the subspace increases, with an inflection at 39 dimensions (out of 120 initially) according to the finite differences method. As discussed by @Runghen2021Exploiting, there is often a trade-off between the number of dimensions to use (more dimensions are more computationally demanding) and the quality of the representation. In panel **B**, we show the increase in cumulative variance explained at each rank, and visualize that using 39 ranks explains about 70% of the variance in the empirical metaweb. This provides different information from the $L_2$ loss (which is averaged across interactions), as it works on the eigenvalues of the embedding, and therefore captures higher-level features of the network. In panel **C**, we show positions of hosts and parasites on the first two dimensions of the left and right subspaces. Note that these values largely skew negative, because the first dimensions capture the coarse structure of the network: most pairs of species do not interact, and therefore have negative values. Finally in panel **D**, we show the predicted weight (*i.e.* the result of the multiplication of the RDGP subspaces at a rank of 39) as a function of whether the interactions are observed, not-observed, or unknown due to lack of co-occurrence in the original dataset. This reveals that the observed interactions have higher predicted weights, although there is some overlap; the usual approach to identify potential interactions based on this information would be a thresholding analysis, which is outside the scope of this manuscript (and is done in the papers cited in this illustration). Because the values returned from RDPG are not bound to the unit interval, we performed a clamping of the weights to the unit space, showing a one-inflation in documented interactions, and a zero-inflation in other species pairs. Panel **D** specifically shows that species pairs with no documented co-occurrence have weights that are not distinguishable from species pairs with no documented interactions; in other words, looking at the embedding, species that do not co-occur are not easily distinguished from species that do not interact. This suggests that (as befits a host-parasite model) the ability to interact is a strong predictor of co-occurrence.

@@

In @fig-illustration-2, we relate the values of latent variables for hosts to different ecologically-relevant data; we can perform this additional step, because the results presented in @fig-illustration-1 show that we can extract an embedding of the metaweb that captures enough variance to be relevant. Importantly, this is true for both $L_2$ loss (indicating that RDPG is able to capture pairwise processes) and the cumulative variance explained (indicating that RDPG is able to capture network-level structure), which suggests that these approaches may allow to predict interactions *and* network structure. In panel **A**, we show that host with a higher value on the first dimension have fewer parasites. This relates to the body size of hosts in the *PanTHERIA* database [@Jones2009Pantheria], as shown in panel **B**: interestingly, the position on the first axis is only weakly correlated to body mass of the host; this matches well established results showing that body size/mass is not always a direct predictor of parasite richness in terrestrial mammals [@Morand1998Density], a result we observe in panel **C**. Finally, in panel **D**, we can see how different taxonomic families occupy different positions on the first axis, with *e.g.* Sciuridae being biased towards higher values. These results show how we can look for ecological informations in the output of network embeddings, which can further be refined into the selection of predictors for transfer learning.

@@

Metaweb inference seeks to provide information about the interactions between species at a large spatial scale, typically a scale large enough to be considered of biogeographic relevance (indeed, many of the examples covered in the introduction span areas larger than a country, some of them global). But as @Herbert1965Dune rightfully pointed out, "\[y\]ou can't draw neat lines around planet-wide problems"; any inference of a metaweb must therefore contend with several novel, interwoven, families of problems. In this section, we outline three that we think are particularly important, and discuss how they may be addressed with subsequent data analysis or simulations, and how they emerge in the specific context of using embeddings; some of these issues are related to the application of these methods at the science-policy interface. Adressing these considerations as part of the methodological discussion is particularly important, as the construction of metawebs can perpetuate legacies of biases in data (Box 2).
